---
title: "Assignment 2"
author: "Otgonzaya Battulga"
date: "2025-03-30"
format: html
editor: visual
---
# Building a Prediction Model on House Prices.

In this report, I have developed a pricing model to support the operation of a chain of Airbnb properties by building and evaluating machine learning models that predict listing prices based on property characteristics, host attributes, amenities, and location.

The core dataset used is the **Q3 2024 Airbnb listings for Florida**, sourced from **Inside Airbnb**. This dataset contains over 16,000 observations, providing a comprehensive snapshot of the Florida short-term rental market. Extensive data wrangling was performed, including the extraction of key amenities, handling of missing values, and transformation of categorical variables into model ready features.

**Five predictive models were developed and compared:**

- Ordinary Least Squares (OLS)
- LASSO Regression
- Random Forest
- XGBoost
- CART

Model performance was evaluated based on predictive accuracy (RMSE and R^2) and training time, summarized in a horserace table. The best-performing models, Random Forest and XGBoost were further analyzed to explore feature importance and identify key price drivers.

**To test the validity and generalizability of the models, two additional datasets were introduced:**

- A later snapshot from the same region: Florida Q4 2024
- A different market within the U.S.: Denver Q4 2024

These datasets were used to evaluate how well the models perform across time and geographic location, providing insights into their robustness and potential for wider deployment across an Airbnb portfolio.

```{r loading libraries}
#| output: false
#| echo: false
#| message: false
#| warning: false

library(readxl)
library(tidyverse)
library(caret)
library(glmnet)
library(randomForest)
library(xgboost)
library(e1071)
library(Metrics)
library(rpart)
library(rpart.plot)
library(dplyr)
library(ggplot2)
library(knitr)
library(MetBrewer)
library(scales)
library(kableExtra)
```

```{r importing the dataset}
#| echo: false
#| warning: false

# Set the path to your dataset
file_path <- "data/florida-q3-listings.xlsx"

# Read the Excel file
df <- read_excel(file_path)
```

```{r see column names}
#| output: false
#| echo: false

names(df)
```

